---
title: "Cirrhosis Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Cirrhosis Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/fedesoriano/cirrhosis-prediction-dataset\>*

### Reference:

*\<fedesoriano. (August 2021). Cirrhosis Prediction Dataset. Retrieved [Date Retrieved] from https://www.kaggle.com/fedesoriano/cirrhosis-prediction-dataset.\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Exploratory Data Analysis
## Load dataset
```{r load dataset}
# Load dataset
liver_data <- read.csv("cirrhosis.csv", colClasses = c(
  N_Days = "numeric",
  Status = "factor",
  Drug = "factor",
  Age = "numeric",
  Sex = "factor",
  Ascites = "factor",
  Hepatomegaly = "factor",
  Spiders = "factor",
  Edema = "factor",
  Bilirubin = "numeric",
  Cholesterol = "numeric",
  Albumin = "numeric",
  Copper = "numeric",
  Alk_Phos = "numeric",
  SGOT = "numeric",
  Tryglicerides = "numeric",
  Platelets = "numeric",
  Prothrombin = "numeric",
  Stage = "factor"
))

# Display the structure of the dataset
str(liver_data)

# View the first few rows of the dataset
head(liver_data)

# View the dataset in a separate viewer window
View(liver_data)
```

## Measures of Frequency
```{r Measures of Frequency}
# Frequency of categorical variables
table(liver_data$Status)
table(liver_data$Drug)
table(liver_data$Sex)
table(liver_data$Ascites)
table(liver_data$Hepatomegaly)
table(liver_data$Spiders)
table(liver_data$Edema)
table(liver_data$Stage)
```

## Measures of Central Tendency
```{r MOCT}
# Mean, median, and mode for numerical variables
mean(liver_data$Age)
median(liver_data$Age)
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}
Mode(liver_data$Age)

mean(liver_data$Bilirubin)
median(liver_data$Bilirubin)
Mode(liver_data$Bilirubin)
```

## Measures of Distribution
```{r MOD}
# Range for numerical variables
range(liver_data$Age)
range(liver_data$Bilirubin)

# Variance and standard deviation
var(liver_data$Age)
sd(liver_data$Age)
var(liver_data$Bilirubin)
sd(liver_data$Bilirubin)

# Quartiles
quantile(liver_data$Age)
quantile(liver_data$Bilirubin)
```

## Measures of Relationship
```{r MOR}
# Correlation matrix
cor_matrix <- cor(liver_data[, c("Age", "Bilirubin", "Cholesterol", "Albumin", "Copper", "Alk_Phos", "SGOT", "Tryglicerides", "Platelets", "Prothrombin")])
print(cor_matrix)
```

## ANOVA
```{r ANOVA}
# One-Way ANOVA example (e.g., comparing Age across different Status)
anova_age <- aov(Age ~ Status, data = liver_data)
summary(anova_age)

# Two-Way ANOVA example (e.g., comparing Age across Sex and Drug)
anova_age_sex_drug <- aov(Age ~ Sex + Drug, data = liver_data)
summary(anova_age_sex_drug)
```

## Plots
```{r Plots}
# Histogram for Age
hist(liver_data$Age, main = "Histogram of Age", xlab = "Age")

# Bar plot for Status
barplot(table(liver_data$Status), main = "Bar Plot of Status")

# Scatter plot for Age vs. Bilirubin
plot(liver_data$Age, liver_data$Bilirubin, main = "Scatter Plot of Age vs. Bilirubin", 
     xlab = "Age", ylab = "Bilirubin")

# Box plot for Age by Status
boxplot(Age ~ Status, data = liver_data, main = "Box Plot of Age by Status", 
        xlab = "Status", ylab = "Age")

```

# Preprocessing and Data Transformation
## Missing Values
```{r missing values}
# Check for missing values in the dataset
missing_values <- colSums(is.na(liver_data))

# Print columns with missing values
print(missing_values[missing_values > 0])
```

## Clean Dataset
```{r Clean dataset}
# Delete rows with missing values
cleaned_liver_data <- na.omit(liver_data)

# Check the dimensions of the cleaned dataset
dim(cleaned_liver_data)

# Check for missing values in the cleaned dataset
missing_values_cleaned <- colSums(is.na(cleaned_liver_data))

# Print columns with missing values (if any)
print(missing_values_cleaned[missing_values_cleaned > 0])
```

# Training Model
## Data Splitting
```{r Data Splitting}
# Set seed for reproducibility
set.seed(123)

# Split data into training and testing sets (e.g., 80% train, 20% test)
train_indices <- sample(1:nrow(cleaned_liver_data), 0.8*nrow(cleaned_liver_data))
train_data <- cleaned_liver_data[train_indices, ]
test_data <- cleaned_liver_data[-train_indices, ]

dim(train_data)
dim(test_data)
```

## Bootstrapping
```{r Bootstrapping}
# Perform bootstrapping on Cholesterol
bootstrap_chol <- function(data, num_bootstraps) {
  bootstrap_means <- numeric(num_bootstraps)
  for (i in 1:num_bootstraps) {
    bootstrap_sample <- sample(data$Cholesterol, replace = TRUE)
    bootstrap_means[i] <- mean(bootstrap_sample)
  }
  return(bootstrap_means)
}

# Number of bootstraps
num_bootstraps <- 1000

# Run bootstrapping
bootstrap_means_chol <- bootstrap_chol(cleaned_liver_data, num_bootstraps)

# Summary of bootstrapped means
summary(bootstrap_means_chol)
```

## Cross-validation
```{r Cross-validation}
# Example using basic k-fold cross-validation with logistic regression
library(caret)

# Define control parameters for cross-validation
control <- trainControl(method = "cv", number = 5)


# Train logistic regression model using cross-validation
model <- train(Status ~ ., data = train_data, method = "rf", trControl = control)

# Print model performance
print(model)
```

## Training Different Models
```{r Training Different Models}
# Train multinomial logistic regression model
library(nnet)
model_multinom <- multinom(Status ~ ., data = train_data)

# Print model summary
summary(model_multinom)

# Make predictions on the test set
predictions_multinom <- predict(model_multinom, newdata = test_data)

# Evaluate model performance
confusionMatrix(predictions_multinom, test_data$Status)

# Train random forest model
library(randomForest)
model_rf <- randomForest(Status ~ ., data = train_data)

# Print model summary
print(model_rf)

# Make predictions on the test set
predictions_rf <- predict(model_rf, newdata = test_data)

# Evaluate model performance
confusionMatrix(predictions_rf, test_data$Status)

# Train SVM model
library(e1071)
model_svm <- svm(Status ~ ., data = train_data)

# Print model summary
print(model_svm)

# Make predictions on the test set
predictions_svm <- predict(model_svm, newdata = test_data)

# Evaluate model performance
confusionMatrix(predictions_svm, test_data$Status)
```

## Performance Comparison
```{r Performance Comparison}
# Load necessary libraries
library(caret)
library(e1071)  # For SVM
library(nnet)  # For neural networks
library(randomForest)  # For random forest

# Define control for cross-validation
control <- trainControl(method = "cv", number = 10)

# Train SVM model
set.seed(123)  # For reproducibility
model_svm <- train(Status ~ ., data = train_data, method = "svmRadial", trControl = control)

# Train neural network (nnet) model
set.seed(123)
model_nnet <- train(Status ~ ., data = train_data, method = "nnet", trControl = control, trace = FALSE, linout = TRUE)

# Train multinomial logistic regression model
set.seed(123)
model_multinom <- train(Status ~ ., data = train_data, method = "multinom", trControl = control, trace = FALSE)

# Train random forest model
set.seed(123)
model_rf <- train(Status ~ ., data = train_data, method = "rf", trControl = control)

# Aggregate results using resamples
results <- resamples(list(SVM = model_svm, NNet = model_nnet, Multinom = model_multinom, RF = model_rf))

# Print summary of the results
summary(results)

# Plot comparisons
dotplot(results)
```

## Saving Model
```{r Saving Model}
# Load the saved SVM model
loaded_model_svm <- readRDS("./models/svm_model.rds")

# Prepare new data for prediction
new_data <- data.frame(
  N_Days = 4000,
  Age = 5000,
  Sex = "M",
  Drug = "D-penicillamine",
  Status = "C",  # You can put any placeholder value here since it's the target variable
  Ascites = "N",
  Hepatomegaly = "N",
  Spiders = "N",
  Edema = "N",
  Bilirubin = 1.2,
  Cholesterol = 150,
  Albumin = 3.5,
  Copper = 50,
  Alk_Phos = 100,
  SGOT = 40,
  Tryglicerides = 120,
  Platelets = 200,
  Prothrombin = 12,
  Stage = "1"  # You can put any placeholder value here since it's not needed for prediction
)

# Use the loaded model to make predictions for new data
predictions_loaded_model <- predict(loaded_model_svm, newdata = new_data)

# Print predictions
print(predictions_loaded_model)

```

